---
title: "Writing query functions in neuprintr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{customquery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

neuPrint includes both an [API](https://neuprint.janelia.org/help/api) which
provides a range of queries as well as the option to send *custom queries* 
written in the [Cypher query language](https://neo4j.com/developer/cypher-query-language/)
of the [Neo4j](https://neo4j.com) graph database.

It is probably worth making queries via the API if they will solve your problem.
However custom queries offer maximum flexibility.

## Do you need a function?

It is a great idea to wrap your queries in a function. This will make 
your code cleaner and more reusable. But do check the documentation to ensure
that there isn't already a neuprintr function that does the job. If there is
something that almost looks correct, then consider asking us to adapt it!

## Custom query functions

We'll start by giving a basic example query function. Most of the time you should
just be able to edit this example without worrying about the details. The
subsequent sections give you some information about those details.

### Basic example

This function takes one or more bodyids as input and returns the name of the 
neurons.

```{r, eval=FALSE}
neuprint_get_neuron_names <- function(bodyids, dataset = NULL, all_segments = FALSE, conn = NULL, ...) {
  all_segments.json = ifelse(all_segments,"Segment", "Neuron")
  cypher = sprintf("WITH %s AS bodyIds UNWIND bodyIds AS bodyId MATCH (n:`%s`) WHERE n.bodyId=bodyId RETURN n.instance AS name",
                   id2json(bodyids),
                   all_segments.json)

  nc = neuprint_fetch_custom(cypher=cypher, conn = conn, dataset = dataset, ...)
  d =  unlist(lapply(nc$data,nullToNA))
  names(d) = bodyids
  d
}
```

`bodyids` is the main input argument. These typically come in as either character
vector or numeric format. The internal function `id2json()` will look
after formatting them appropriately for the query.

In this function, `cypher` is the actual query written in the 
[Cypher query language](https://neo4j.com/developer/cypher-query-language/). 
One helpful tip. *You can press the i key in neuPrint explorer to reveal the cypher query*!

Many functions that operate on neurons will have an argument controlling whether
they operate only for larger objects (aka *Neuron*) or also on fragments (aka *Segment*).
Restricting queries to *Neuron* can result in big speed-ups in some cases.
You can provide an `all_segments` argument to handle this option.

Finally the results that come back from `neuprint_fetch_custom` are typically
in a big list object. For simple results, you can just `unlist()` this to make a
vector. In this case a function `nullToNA` is first applied in order to ensure
that any values that come back as `NULL` are converted to `NA`; this is necessary
to ensure that you can make a vector - vector objects can only contain `NA`s not `NULL`s.

### Preparing the query

As you can see this step uses the `sprintf` function to interpolate variables
into a string. You could do this in other ways (e.g. the `paste()` function or
the `glue()` package). You may need to watch out for quoting issues if you should
need to use single or double quotes inside your queries.

### Body ids

You need to be a little careful with the handling of body ids. These are 64 bit
integers (often called `bigint`s), which are commonly used as keys in databases.
However like many programming languages R does not have a native 64 bit integer
type. R's default numeric format is double width floating point. This can 
exactly handle numbers with up to 53 bits of precision. 

However, the biggest integer that we need to worry about is 2^64-1 = 9223372036854775806. This cannot be represented as a numeric. I have yet to 
spot a body id in this upper range, but the `id2json` function will take care
that you do not use one. If you need to specify a large bodyid as input to your
function then you should insist on 

* character vector input
* 64 bit integers provided by `bit64::integer64()`

The latter is arguably the best approach, but it relies on an add-on package
(bit64) rather than base R.

Bottom line

* always use `id2json()` on your body ids when building your cypher
* if you work with your body ids inside your function then do `id2char` to put
them into the most robust standard form (a character vector).

`id2bit64` is also available. This sounds better, but these `bit64::integer64()`
are a bit fragile. In particular you cannot turn them into a list or run `lapply`
without losing their class.

### Connecting to neuPrint

In order to make a query, you need to specify which neuPrint *server* you want to
talk to as well as the *dataset* you would like to use. The server is specified
by a `neuprint_connection()` object. 99% of the time you will not need to do 
anything as this will be handled transparently by a one time user setting to
specify their preferred server and authentication token. The same is true of the dataset parameter.

However your function must allow people to specify both of these things if they
wish. Therefore your function should look something like this in outline:

```{r, eval=FALSE}

myquery <- function(query, conn=NULL, dataset=NULL, ...) {
  neuprint_fetch_custom(query, conn=conn, dataset=dataset, ...)
}

```

Internally `neuprint_fetch_custom()`, which will look after the `dataset` 
argument, and then go on to call the low level `neuprint_fetch()`, which will
ensure that the connection object is valid (or use `neuprint_login()` to make one).

### Processing the results

The sample function just ran `unlist()` on the list returned by `neuprint_fetch_custom()`. 
This is fine in many cases. You can also pass the `simplifyVector` argument to
`neuprint_fetch_custom()` and this will clean up many forms of list. See 
`jsonlite::fromJSON()` for details of how this operates.

If you are getting a nested list specifying
a table (i.e. a `data.frame` or spreadsheet like result) then there is a
function `neuprint_list2df()` which will do a lot of the heavy lifting.
It is strongly recommended to make use of this whenever possible. See
`neuprint_get_meta()` for an example.

## Need help?

Feel free to ask for help on the [nat-user google group](https://groups.google.com/forum/#!forum/nat-user) or by 
[making an issue](https://github.com/natverse/neuprintr/issues/).
